---
title: "Maximum likelihood: *Tsuga* mortality"
date: "02.02.2026"
output: 
  html_document:
    self_contained: false
    lib_dir: lib
    css: ex.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. Load the tsuga dataset from this repository.

```{r}
library(data.table)
## working directory should be: vu_advstats_students
tsuga = readRDS("data/tsuga.rds")

## if you don't have the repository saved locally
# tsuga = readRDS(url("https://github.com/ltalluto/vu_advstats_students/raw/main/data/tsuga.rds"))
```

This dataset gives statistics about *Tsuga canadensis* observed over multiple years in forest plots in eastern Canada. Included are the number of trees `born`, the number observed to have `died` that year, the total number of trees (including dead ones) `n`, and the climate. Filter the dataset so that it contains only observations from the year 2005 with at least 1 individual (`n > 0`)

   a. Write a function taking three parameters:
      i. `p`: a *single value*, the probability that a randomly chosen individual is dead
      ii. `n`: a *vector*, the number of trees in each plot
      iii. `k`: a *vector*, the number of dead trees in each plot
      iv. The function should return the *log likelihood*: $\log pr(n,k|p)$

```{r}
# n and k are vectors
lfun = function(p, n, k) {
   ## function body
   return() ## a single value!
}
```

   
   b. Plot the log likelihood across various values of `p`
		* **Hint**: you will need to run `lfun` once for each value of `p`. This is most efficiently accomplished using `sapply`, but a `for` loop will also work.
   c. Use `optim` to find the MLE for `p`
   d. Is the answer different from `mean(dat$died/dat$n)`? If so, why?
   e. Write two more functions, one to estimate a prior for *p*, and one to compute the log posterior. You may choose the prior distribution and hyperparameters as you like, but they should respect the constraint that *p* must be between zero and one.
   f. Plot the prior and unnormalized posterior. Compare plots with different prior hyperparameters
   g. Compute the maximum a posteriori estimate for *p* using `optim`.
   h. Write a stan model to compute the maximum a posteriori estimate for p, and then fit it using `optimizing`.

### Bonus

2. Write a Stan program to estimate the mean **number of trees per plot.**
    a. What kind of **statistical process** could generate these data?
    b. What is an appropriate likelihood function? prior?
    c. Compute the MAP estimate.
    d. How does the MAP compare to `mean(dat$n)`?

3. Use the MAP estimate from 2c and the `r****()` function corresponding to the likelihood function to simulate a new dataset, with as many observations as in the original data.
   a. Compare a histogram and summary statistics of your simulated data with the real data. Note any similarities and differences.
   b. Consider how you could improve this **generative model** to better describe the dataset. Do you need a new likelihood function?

