---
title: "Natal vs. Breeding Dispersal in Birds"
date: "27.11.2023"
output: 
  html_document:
    self_contained: false
    lib_dir: lib
    css: ex.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will use a dataset (from [Fandos et al. 2022](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2656.13838)) of dispersal distances of European birds. One important question is whether birds overall have greater dispersal requirements when first leaving the nest where they hatched (**natal dispersal**) or when dispersing as adults among different breeding sites (**breeding dispersal**).

The code below will download the data, and reshape it into a form that is useful for us.

```{r message = FALSE, cache = TRUE, eval = FALSE}
library(data.table)


url = "https://zenodo.org/records/7191344/files/Table_S14_%20species_dispersal_distances_v1_0_2.csv?download=1"
disp = fread(url)

# get rid of columns we won't use, and subset to only breeding/natal dispersal
disp = disp[type %in% c("breeding", "natal"), 
			.(species, median, n, function_id, function_comparison, type, sex_code)]

# they fit four dispersal functions per species/type/sex
# the column function_comparison tells you how good each fit was relative to the others
# we will use it to compute the weighted mean dispersal distance across the different models
disp = disp[, .(disp_dist = weighted.mean(median, function_comparison, na.rm - TRUE), 
				n = sum(n, na.rm = TRUE)), by = .(species, type, sex_code)]

# we will further aggregate by sex (since the paper found little difference among sexes)
# this time with sample size as the weights
disp = disp[, .(disp_dist = weighted.mean(disp_dist, n, na.rm = TRUE)), by = .(species, type)]

# use dcast to put the dispersal types next to each other
disp = dcast(disp, species ~ type, value.var = "disp_dist")

# and drop all cases where we don't have both types
disp = disp[complete.cases(disp)]
```

The original paper details many important factors that might influence dispersal distance, but we will focus on a relatively simple hypothesis: **Across all species, natal dispersal exceeds breeding dispersal**.

Some guidance to get you thinking about the exercise:

1. Make some plots exploring the hypothesis.
2. You probably have an idea of a basic frequentist test for this hypothesis. Go ahead and do it. What's the result? Do the data fit the assumptions of the test?
3. Can you design a **Bayesian model** in Stan that is both appropriate for the data (including making reasonable distributional assumptions) and models the hypothesis you want to test? Try it out, using the tools you know:
	a. Think in terms of the statistical process generating the data (if your hypothesis is true).
	b. Think of **parameters** that can stand in for your hypothesis.
	c. Graph the model, and write equations.
	d. Translate the graph into Stan code and try to fit the model.

