---
title: "Spatial Model Implementation"
author: "Lauren Talluto"
date: "14.05.2021"
output:
  slidy_presentation:
    theme: cerulean
    toc_depth: 2
    css: rmd_style.css
---

```{r setup, include=FALSE, results = "hide"}
knitr::opts_chunk$set(fig.width=5.5, fig.height=5.5, collapse = TRUE, comment = "##", dev="png")
cols = c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF")

library(data.table)
library(ggplot2)
library(gridExtra)
library(rstan)
library(bayesplot)
library(loo)
```

## *Tsuga* mortality revisited

<div class="left lt">
* Here using a subset of data from 2005, considering only *Tsuga* again (to keep our model simpler)
* There are clear spatial patterns!

```{r}
tsuga = readRDS("exercises/data/tsuga_spatial.rds")
pl1 = ggplot(tsuga, aes(x=X, y=Y, colour = died/n)) + geom_point() + 
	coord_fixed() + theme_minimal() + labs(color="pr(death)") + 
	scico::scale_color_scico(palette="imola") + 
	ylab("Latitude (m)") + xlab("") + coord_fixed()
pl2 = ggplot(tsuga, aes(x=X, y=Y, colour = annual_mean_temp)) + 
	geom_point() + coord_fixed() + theme_minimal() + 
	scico::scale_color_scico(palette="vik") + 
	labs(color="temperature") + ylab("Latitude (m)") + 
	xlab("Longitude (m)") + coord_fixed()

```

</div>
<div class="right rt">
```{r echo = FALSE, fig.width = 9}
grid.arrange(pl1, pl2, nrow=2)

```

</div>

## Fitting the model
<div class="left lt">
* We compare 4 models
   - Spatial + temperature effect
   - Spatial only
   - Temperature only
   - Spatial + temperature + precipitation

**Note**: To run the code in this .rmd, you first need to run `r/tsuga_fit.r`, with your working directory set to `docs`.

```{r cache=TRUE}
models = readRDS("../scratch/gp_models.rds")
p1 = mcmc_trace(as.matrix(models[[1]], pars=c("a","b"))) + 
	theme_minimal() + ggtitle("T")
p2 = mcmc_trace(as.matrix(models[[2]], pars=c("a","rho"))) + 
	theme_minimal() + ggtitle("GP")
p3 = mcmc_trace(as.matrix(models[[3]], pars=c("a","rho"))) + 
	theme_minimal() + ggtitle("GP+T")
p4 = mcmc_trace(as.matrix(models[[4]], pars=c("a","rho"))) + 
	theme_minimal() + ggtitle("GP+T+P")

```

</div>
<div class="right rt">
```{r echo=FALSE, fig.height=9.5, fig.width=7}
grid.arrange(p1, p2, p3, p4, nrow=4)

```

</div>

## Model Comparison w/ loo

<div class="left lt">
* Loo has a hard time with the GP models

```{r cache=TRUE}
loos = readRDS("../scratch/gp_loos.rds")
print(loos[[2]])
```


* WAIC is also problematic

```{r cache=TRUE}
waics = lapply(models, function(x) waic(as.matrix(x, pars="loglik")))
```

</div>
<div class="right rt">

* At this point we must proceed with caution.
   - Try to reparameterize/transform
   - Proceed with inference from a single model
   - Biggest model is safest, it includes everything
* Model averaging is not an option

```{r}
loo_compare(loos)
```

```{r}
loo_compare(waics)
```

</div>

## Visualising the scale of the spatial effect

```{r, cache=TRUE, echo=FALSE, warning=FALSE, fig.width=12}
kern = function(alpha, rho, dist)
	alpha^2 * exp(-(dist^2)/(2*rho^2))

kern_summary = function(samples, dist) {
	## we use a for loop here to reduce memory usage
	res = matrix(NA, nrow=length(dist), ncol=3, dimnames=list(NULL, c("median", "lower", "upper")))
	for(i in 1:nrow(res)) {
		kvals = mapply(kern, alpha = samples[,'alpha'], rho = samples[,'rho'], MoreArgs= list(dist = dist[i]))
		res[i,] = quantile(kvals, c(0.5, 0.05, 0.95))
	}
	res
}

standat = readRDS("../scratch/standat.rds")
kern_samps = lapply(models[2:4], as.matrix, pars = c("alpha", "rho"))

dd = seq(0, max(dist(standat$coords)), length.out=400)
ypl = parallel::mclapply(kern_samps, kern_summary, dist=dd, mc.cores=3)

# the original distances were on scales of 100 km, so we convert back here
# so we can plot covariance against distance in km
ypl2 = lapply(ypl, function(x) {
	x = as.data.table(x)
	x$distance = dd*100
	x
})
ypl2 = rbindlist(ypl2, idcol="model")

ggplot(ypl2, aes(x = distance, y = median, colour=model)) + geom_line() + geom_ribbon(aes(ymin=lower, ymax=upper, fill=model), alpha=0.5) + 
	theme_minimal() + xlab("Distance (km)") + ylab("Covariance") + facet_grid(~model) + xlim(0, 500)

```


## Comparing parameter estimates

<div class="left lt">

```{r echo=FALSE, warning=FALSE, fig.width=9, fig.height=9}
samples = lapply(models[c(1,3,4)], as.matrix, pars=c("a", "b"))
samples_comb = rbindlist(lapply(samples, data.table), idcol="model", fill=TRUE)
samples_comb = melt(samples_comb, id.vars = "model", variable.name = "parameter")
samples_comb[parameter == "b[1]", parameter := "temperature"]
samples_comb[parameter == "b[2]", parameter := "precipitation"]

# stats = lapply(samples, function(x)
# 	data.table(param = c("a", "b"), median = apply(x, 2, median), lower = apply(x, 2, quantile, 0.05), upper = apply(x, 2, quantile, 0.95)))
# stats = rbindlist(stats, idcol="model")
ggplot(samples_comb, aes(x=parameter, y=value, fill=model)) + 
	geom_boxplot(notch=TRUE, outlier.shape=NA) + theme_minimal() + ylim(-3, 2.5)
```

</div>

<div class="right rt">


```{r echo=FALSE, warning=FALSE, figure.width=9, figure.height=9}
mcmc_pairs(as.matrix(models[[3]], pars=c("a", "b", "alpha", "rho")))
```

</div>


## Temperature response curves
<div class="left lt">

* Note that the prediction for any data point might vary considerably from this
* This shows the effect after marginalising out any spatial effects

```{r echo=FALSE, warning=FALSE, fig.width=9, fig.height=9}
xx = seq(min(standat$x[,'annual_mean_temp']), max(standat$x[,'annual_mean_temp']), length.out=250)
prep_rc = function(mod) {
	samps_params = as.matrix(mod, pars=c('a', 'b'))
	xx = rbind(1, xx) ## add intercept
	if(ncol(samps_params) == 3)
		xx = rbind(xx, 0)  ## add zero effect for precip; set to median
	mu = plogis(samps_params %*% xx)
	data.table(
		temperature = xx[2,] * sd(tsuga$annual_mean_temp) + mean(tsuga$annual_mean_temp),
		median = apply(mu, 2, median),
		lower = apply(mu, 2, quantile, 0.05),
		upper = apply(mu, 2, quantile, 0.95))
}
resp_curve = rbindlist(lapply(models[c(1,3,4)], prep_rc), idcol="model")

ggplot(resp_curve, aes(x=temperature, y = median, colour=model)) + geom_line(size=1.2) + 
	geom_ribbon(aes(x=temperature, ymin=lower, ymax=upper, fill=model), alpha=0.4, size=0) + theme_minimal() +
	xlab("Mean Annual Temperature") + ylab("Mortality Probability") + 
	ggtitle("Predictions without Spatial Effects")


```

</div>

<div class="right rt">

```{r echo=FALSE, warning=FALSE, fig.width=9, fig.height=9}
plpts = rbindlist(lapply(models, function(mod) {
	samps = as.matrix(mod, pars="mu")
	data.table(temperature = tsuga$annual_mean_temp,
		median = apply(samps, 2, median),
		lower = apply(samps, 2, quantile, 0.05),
		upper = apply(samps, 2, quantile, 0.95))
}), idcol="model")
ggplot(plpts, aes(x=temperature, y=median, colour=model)) + geom_point() + theme_minimal() +
	xlab("Mean Annual Temperature") + ylab("Mortality Probability") + facet_wrap(~model) + 
	ggtitle("Predictions w/ Spatial Effects")

```
</div>

