---
title: "Distribution Exercises"
date: "02.05.2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=4)
```

1. In addition to the `d` and `p` functions mentioned in class, R includes `q` and `r` functions. What do these do? Explore the help files.
**`q`: quantiles; returns the inverse of `p`. So `qnorm(0.5)` gives you the value of a normal distribution such that the probability of observing a smaller value is 0.5.**

**`r`: provide random number generation from a distribution**

2. Austria had a birth rate of approximately 9.9 births per 1000 people in 2019. Assume this rate is constant over the year, and that all births are independent.
   a. For a randomly selected cohort of 1000 people, what is the probability of observing exactly 11 births over a 1-year period?
   
**You can model this as a binomial process with $n=1000,p=0.0099$, or Poisson, with $\lambda = 9.9$.**

```{r}
c(binom = dbinom(11, 1000, 0.0099),
pois = dpois(11, 9.9))
```

   
   b. What about observing **11 or more** births?

**Lots of ways to arrive at the same answer**

```{r}
# the same logic here applies for binom

c(
   ## ppois(10, ...) is the prob of 10 or less events
   ## 1 - ppois(10, ...) gives the prob of more than 10
   1 - ppois(10, 9.9),  
   ## we can achieve the same thing with the lower.tail argument
   ppois(10, 9.9, lower.tail=FALSE),
   ## the poisson is discrete, so we can also sum the mass function
   1 - sum(dpois(0:10, 9.9)),
   ## technically we need to go to infinity, but the error is small
   sum(dpois(11:1e6, 9.9))
)
```


3. Is a **probability density** the same as a **probability**?
   a. Human height is normally distributed within populations. From 1980-1994 within 20 wealthy countries, mean female height was 164.7 cm, with a standard deviation of 7.1 cm. What is the **maximum probability density** of this normal distribution, and what is the x-value $x_{max}$ at which maximum probability density occurs?

**The maximum density of the normal is at the mean, so $x_{max}=164.7$** 

```{r}
xmax = 164.7
mean_ht = 164.7
sd_ht = 7.1
dnorm(xmax, mean = mean_ht, sd = sd_ht)
```

   b. What is the probability that a female in this time period has a height exactly equal to $x_{max}$
   c. If the maximum probability density and the $pr(x_{max})$ are not the same, why not? Do these answers make sense?
   
**The probability is zero, because $x$ is continuous. We computed the density in part a. We can compute the actual probability (probability mass) by integrating; clearly it will be zero, because the integral has a width of zero.**

```{r}
integrate(dnorm, lower = xmax, upper = xmax, mean = mean_ht, sd = sd_ht)
```

   d. What is the probability that a woman has a height in the range $x_{max} \pm 3$

**There is nonzero mass between two different values. We can either integrate as before, or use pnorm**
   
```{r}
c(pnorm = pnorm(xmax + 3, mean_ht, sd_ht) - pnorm(xmax - 3, mean_ht, sd_ht),
integrate = integrate(dnorm, lower = xmax - 3, upper = xmax + 3, mean = mean_ht, sd = sd_ht)["value"])
```
   
4. For the same distribution, what is the 40th percentile for height? In other words, what is the value $x$ such that the probability of observing x or less is 0.4? --- $pr(X \le x) = 0.4$

```{r}
qnorm(0.4, 164.7, 7.1)
```

   a. What is $x$ if $pr(X > x) = 0.4$?

```{r}
qnorm(0.4, 164.7, 7.1, lower.tail=FALSE)
qnorm(0.6, 164.7, 7.1, lower.tail=TRUE)
```
