---
title: "Distribution Exercises"
date: "03.11.2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=4)
```

1. In addition to the `d` and `p` functions mentioned in class, R includes `q` and `r` functions. What do these do? Explore the help files.
**`q`: quantiles; returns the inverse of `p`. So `qnorm(0.5)` gives you the value of a normal distribution such that the probability of observing a smaller value is 0.5.**

**`r`: provide random number generation from a distribution**

2. Austria had a birth rate of approximately 9.9 births per 1000 people in 2019. Assume this rate is constant over the year, and that all births are independent.
   a. For a randomly selected cohort of 1000 people, what is the probability of observing exactly 11 births over a 1-year period?
   
**You can model this as a binomial process with $n=1000,p=0.0099$, or Poisson, with $\lambda = 9.9$.**

```{r}
c(binom = dbinom(11, 1000, 0.0099),
pois = dpois(11, 9.9))
```

   
   b. What about observing **11 or more** births?

```{r}
# the same logic here applies for binom

1 - ppois(10, 9.9)
ppois(10, 9.9, lower.tail=FALSE)
1 - sum(dpois(0:10, 9.9))
sum(dpois(11:1e6, 9.9)) # technically we need to go to infinity, but the error is small
```


3. Is a **probability density** the same as a **probability**?
   a. Human height is normally distributed within populations. From 1980-1994 within 20 wealthy countries, mean female height was 164.7 cm, with a standard deviation of 7.1 cm. What is the **maximum probability density** of this normal distribution, and what is the x-value $x_{max}$ at which maximum probability density occurs?

**The maximum density of the normal is at the mean, so $x_{max}=164.7$**   

   b. What is the probability that a female in this time period has a height exactly equal to $x_{max}$
   c. If the maximum probability density and the $pr(x_{max})$ are not the same, why not? Do these answers make sense?
   
**The probability is zero, because $x$ is continuous. For the density:**

```{r}
dnorm(164.7, 164.7, 7.1)
```

   d. What is the probability that a woman has a height in the range $x_{max} \pm 3$
   
```{r}
pnorm(164.7+3, 164.7, 7.1) - pnorm(164.7-3, 164.7, 7.1)
integrate(dnorm, lower=164.7-3, upper=164.7+3, mean=164.7, sd=7.1)
```
   
4. For the same distribution, what is the 40th percentile for height? In other words, what is the value $x$ such that the probability of observing x or less is 0.4? --- $pr(X \le x) = 0.4$

```{r}
qnorm(0.4, 164.7, 7.1)
```

   a. What is $x$ if $pr(X > x) = 0.4$?

```{r}
qnorm(0.4, 164.7, 7.1, lower.tail=FALSE)
qnorm(0.6, 164.7, 7.1, lower.tail=TRUE)
```
